---
title: "Group 4 — Youth risk"
subtitle: "Studio wall"
author: "Shivgun, Shreeya and Meghana"
format:
  html:
    toc: true
    toc-depth: 2
---

# Spine

> What do teenagers’ bodies and habits look like, which habits travel together, and where is life long?

This sentence stays at the top. Every panel must attach to it. If a plot cannot be introduced in a clause that hangs off this line, it is off the wall.

# Introduction — the tables we own

This folder is ours for the ten days. Three tables, three shapes.

- **Panel A** (`panel-A.csv`) — YRBSS Quants: height, weight, days active, strength training, hours of sleep.
- **Panel B** (`panel-B.csv`) — YRBSS Quals: helmet, texting while driving, TV, race, gender, grade. Same survey, different columns.
- **Panel C** (`panel-C.csv`) — gapminder 2007 world. Ask a *different* map question than Group 3 (wealth, population symbols, continent).

A and B are the same teenagers cut two ways. Do not mix them into one dishonest figure.

We load from this folder:

```{r}
#| label: setup
library(skimr)
library(mosaic)
library(readr)
library(ggformula) 
##
library(vcd) # (NEW!!) for Panel B
library(broom) # (NEW!!) for clean stat outputs

PanelA <- read.csv(file = "data/panel-A.csv")
PanelB <- read.csv(file = "data/panel-B.csv")
PanelC <- read.csv(file = "data/panel-C.csv")

```

Found tables train the mark. A campus Free Hunch, if we collect one, is a **made** panel of the same pair type — new people, old grammar. It does not replace A, B, or C.

# Panel A — body / measure

*height, weight, physically_active_7d, school_night_hours_sleep are Quant. grade is Qual.*

### 1. See the table

What is a case? Which columns are Quant / Qual? What is missing?

``` r
# glimpse() / inspect() / skim()
```

```{r}
#| label: inspect-panelA-data-1
dplyr::glimpse(PanelA)

```

My PanelA data has:

- 13,583 observations and 08 variables.
- `age`: Quantitative, numeric
- `height`: Quantitative, numeric
- `weight`: Quantitative, numeric
- `physically_active_7d`: Quantitative, numeric
- `strength_training_7d`: Quantitative, numeric
- `gender`: Qualitative, categorical. Encoded <chr>, to be mutated
- `grade`: Qualitative, categorical. Encoded <chr>, to be mutated
- `school_night_hours_sleep`: Qualitative, categorical. Encoded <chr>, to be mutated

```{r}
#| label: inspect-panelA-data-2
skimr::skim(PanelA)
```

Furthermore, on the Qual variables panel-A data has :

- `gender` has 3 levels:
- `grade` has 6 levels:
- `school_night_hours_sleep` has 8 levels:

And on the Quant variables panel-A data has :

- `age` has a mean of 16.15, sd of 1.26, median of 16.00, min of 12.00, and max of 18.00
- `height` has a mean of 1.69, sd of 0.10, median of 1.68, min of 1.27, and max of 2.11
- `weight` has a mean of 67.90, sd of 16.89, median of 64.41, min of 29.94, and max of 180.99
- `physically_active_7d` has a mean of 3.90, sd of 2.56, median of 4.00, min of 0.00, and max of 7.00
- `strength_training_7d` has a mean of 2.94, sd of 2.57, median of 3.00, min of 0.00, and max of 7.00

There is missing data.

```{r}
#| label: mutate-panelA-factors

A_mod <- PanelA |>  mutate( gender = as.factor(gender),
    grade = as.factor(grade),
    school_night_hours_sleep = as.factor(school_night_hours_sleep))

glimpse(A_mod)
    
```

All is well now as the Blue Lanterns say.

### 2. Ask

One question *this* table can answer. One it cannot. Write both.

Looking at the overall Spine statement, I can pose the following Questions:

1.  Which grade has the most students?
2.  What is the gender distribution across the grades?
3.  Do boys strength train more than girls?
4.  Do girls have better school night sleep than boys? **5. Are heavier students more physically active, does it differ between boys and girls?**
5.  How does height differ between boys and girls?
6.  Does grade affect how physically active students are,and does it differ between boys and girls?
7.  Do heavier students (like me) tend to be in higher grades?
8.  Do older students sleep less on school nights, and does it differ between boys and girls? 10.Do students who are more physically active also strength train more?

Possibly cannot answer: 1. Why do some students sleep less than 5 hours on school nights? 2. Does the sleep depend on the professor's/teacher's class they have on the next day? 3. Does their weight depend on the items served in mess/tiffins?

### 3. Choose a mark

Geom, map mark, or (later) test that matches the pair type. Name it before you code.

Histogram Scatter Plot Box Plot Bar graph Density graph

### 4. Predict

What do we think we will see? One sentence, before Run.

### 5. Make

Human brick first. Optional: one AI prompt, pasted below, only for what section 3 named.

1.  Which grade has the most students??

```{r}
#| eval: false
#| echo: true
A_mod %>% gf_bar (~ grade)
```

2.  What is the sex distribution among the grades?

```{r}
#| eval: false
#| echo: true
A_mod %>% gf_bar(~grade, fill = ~gender)

```

3.  Do boys strength train more than girls?

```{r}
#| eval: false
#| echo: true
A_mod %>% gf_histogram(~ strength_training_7d, fill = ~gender, bins = 1, position = "dodge")
```

```{r}
#| eval: false
#| echo: true
A_mod %>% gf_density(~ strength_training_7d, fill = ~gender)
```

4.  Do girls have better school night sleep than boys?

```{r}
#| eval: false
#| echo: true
A_mod %>% gf_bar(~ gender, fill = ~school_night_hours_sleep)
```

**Are heavier students more physically active, does it differ between boys and girls?**

```{r}
A_mod %>% 
  filter(gender %in% c("male", "female")) %>%

  gf_point(weight ~ physically_active_7d, color= ~gender)
```

6.  How does height differ between boys and girls?

```{r}
#| eval: false
#| echo: true
A_mod %>% gf_histogram(~height, fill = ~gender)

```

```{r}
#| eval: false
#| echo: true
A_mod %>% gf_density(~ height, fill = ~gender)
```

7.  Does grade affect how physically active students are, and does it differ between boys and girls?

```{r}
#| eval: false
#| echo: true
A_mod %>%  filter(gender %in% c("male", "female")) %>% 
  gf_boxplot( x = ~grade, y = ~physically_active_7d, fill = ~gender) 
```

8.  Do heavier students (like me) tend to be in higher grades?

```{r}
#| eval: false
#| echo: true
A_mod %>% gf_boxplot( x = ~weight, y = ~grade)

```

9.  Do older students sleep less on school nights, and does it differ between boys and girls?

```{r}
#| eval: false
#| echo: true
A_mod %>%  gf_boxplot(x = ~school_night_hours_sleep, y = ~age, fill = ~gender)
```

10. Do students who are more physically active also strength train more?

```{r}
#| eval: false
#| echo: true
A_mod %>% gf_point( strength_training_7d ~physically_active_7d)

```

### 6. Say

**Weight tells us almost nothing about how many days a student is active.**

Students at every activity level, 0 to 7 days a week, cover nearly the same range of weights, for both boys and girls.

*Heavier students are not visibly more or less active than lighter ones.*

### 7. Limit

This panel can't say anything about fitness or body composition, since weight alone isn't BMI. There is also missing data.

# Panel B — category / talk

*helmet_12m × gender, text_while_driving_30d × grade. Hours of TV may be ordinal — argue before you average it.*

### 1. See the table

What is a case? Which columns are Quant / Qual? What is missing?

``` r
# glimpse() / inspect() / skim()
```

```{r}
#| label: inspect-panelB-data-1
dplyr::glimpse(PanelB)

```

My PanelB data has:

- 13,583 observations and 08 variables.
- `age`: Quantitative, numeric
- `gender`: Qualitative, categorical. Encoded <chr>, to be mutated
- `grade`: Qualitative, categorical. Encoded <chr>, to be mutated
- `hispanic`: Qualitative, categorical. Encoded <chr>, to be mutated
- \`race\`\`: Qualitative, categorical. Encoded <chr>, to be mutated
- `helmet_12m`: Qualitative, categorical. Encoded <chr>, to be mutated
- `text_while_driving_30d`: Qualitative, categorical. Encoded <chr>, to be mutated'
- `hours_tv_per_school_day`: Qualitative, categorical. Encoded <chr>, to be mutated

```{r}
#| label: inspect-panelB-data-2
skimr::skim(PanelB)
```

Furthermore, on the Qual variables panel-A data has :

- `gender` has 3 levels:
- `grade` has 6 levels:
- `hispanic` has 3 levels:
- `race`has 6 levels:
- `helment_12m` has 7 levels:
- `text_while_driving_30d` has 9 levels:
- `hours_tv_per_school_day` has 8 levels:

And on the Quant variables panel-A data has :

- `age` has a mean of 16.15, sd of 1.26, median of 16.00, min of 12.00, and max of 18.00

```{r}
#| label: mutate-panelB-factors

B_mod <- PanelB |>  mutate( gender = as.factor(gender),
    grade = as.factor(grade), hispanic = as.factor(hispanic), race = as.factor(race), helmet_12m = as.factor(helmet_12m), text_while_driving_30d = as.factor(text_while_driving_30d), hours_tv_per_school_day = as.factor(hours_tv_per_school_day)) 

glimpse(B_mod)
```

### 2. Ask

One question *this* table can answer. One it cannot. Write both.

Looking at the overall Spine statement, I can pose the following Questions:

1.  Do boys or girls text while driving more? **2. Do boys or girls skip helmets more?**
2.  Are the students who text while driving older?

Possibly cannot answer:

1.  Why do some students never wear a helmet?
2.  Do students who text while driving have more crashes?

### 3. Choose a mark

Geom, map mark, or (later) test that matches the pair type. Name it before you code.

Bar Graph Box Plot Mosaic Charts

```{r}
#| eval: false
#| echo: true
vcd::structable(grade ~ gender , data = B_mod) %>% vcd::mosaic()
```

```{r}
#| eval: false
#| echo: true
vcd::structable(grade ~ gender , data = B_mod) %>% vcd::mosaic(type = "expected")
```

```{r}
#| eval: false
#| echo: true
mosaic::xchisq.test(grade ~ gender, data = B_mod) 
```

```{r}
#| eval: false
#| echo: true
vcd::structable(gender ~ helmet_12m , data = B_mod) %>% vcd::mosaic()

```

```{r}
#| eval: false
#| echo: true
vcd::structable(gender ~ helmet_12m , data = B_mod) %>% vcd::mosaic(type = "expected")

```

```{r}
#| eval: false
#| echo: true
mosaic::xchisq.test(gender ~ helmet_12m , data = B_mod) 
```

### 4. Predict

What do we think we will see? One sentence, before Run.

### 5. Make

1.  Do boys or girls text while driving more?

```{r}
#| eval: false
#| echo: true
#| 
B_mod %>% gf_bar( ~text_while_driving_30d, fill = ~gender)

```

**2. Do boys or girls skip helmets more?**

```{r}
B_mod %>% gf_bar( ~helmet_12m, fill = ~gender)

```

3.  Are the students who text while driving older?

```{r}
#| eval: false
#| echo: true

B_mod %>% gf_boxplot( x = ~age, y =~text_while_driving_30d)

```

### 6. Say

**Most students never wear a helmet, whether they are boys or girls.**

Never" is the tallest bar by a wide margin, and both boys and girls make up a large part of it.

Helmet use is rare among students of both genders.

### 7. Limit

This panel shows how often students say they wear a helmet, not why they skip it.

The chart shows counts, not proportions, so not an exact boys versus girls comparison.

# Panel C — place

*Same file as Group 3 C. Different question. lifeExp vs gdpPercap, or pop as symbol size.*

Geometry is not in the CSV. Join in R, then use the same seven headings.

``` r
# library(sf)
# library(rnaturalearth)
# world <- ne_countries(scale = "medium", returnclass = "sf")
# # join on country name; list the unmatched labels
```

### 1. See the table

What is a case? Which columns are Quant / Qual? What is missing?

``` r
# glimpse() / inspect() / skim()
```

```{r}
#| label: inspect-panelC-data
dplyr::glimpse(PanelC)

```

### 2. Ask

One question *this* table can answer. One it cannot. Write both.

### 3. Choose a mark

Geom, map mark, or (later) test that matches the pair type. Name it before you code.

### 4. Predict

What do we think we will see? One sentence, before Run.

### 5. Make

Human brick first. Optional: one AI prompt, pasted below, only for what section 3 named.

``` r
# human chunk

# ## Prompt we actually used
# ai_draft
# ours
```

### 6. Say

Caption + one sentence we will stand behind. Title is a sentence, not a column name.

### 7. Limit

What this panel cannot do. Which kit table we refused to abuse.

# Conclusion

What the spine looks like after three panels.

- The question we will defend in ninety seconds
- What we will **not** claim
- Which lines were human, which were machine
- Limits: sample, missingness, the test we did not run
- Gift for the steal-this strip (palette, annotation, join trick)

``` r
# sessionInfo()
```
